Semi-permeable vesicles: mathematical formulation, numerics and physics

### 2020-02-21 vesicle_code/docs/permeableVes/main.tex
finished commenting the initilization routines
started to map out paper a bit better

\documentclass[aps,prl,showpacs]{revtex4}
\usepackage{amsmath,amsfonts,bm, color}
\usepackage{graphicx,epsfig}
%\input vs_macros.tex
%Dfig puts a figure of specified size at the place indicated.
\newcommand{\Dfig}[2]{\epsfig{figure=#1,width=#2cm}}

@@

\title{Semi-permeable vesicles: mathematical formulation, numerics and physics}
\author{}

@@

\section{Problem formulation} \label{sc:formulate}
Consider a semi-permeable vesicle suspended in an unbounded viscous
fluid domain, subjected to an imposed flow $\mbu_\infty(\mbx)$, for any
$\mbx \in \RR^2$.  Assume that the interior and exterior fluids have the same viscosity $\mu$. Let $\mbx$ be the position of the interface $\gamma$, $\mbu$ the fluid velocity and $p$ the pressure. In the vanishing Reynolds number limit, the governing equations for the ambient fluid can be written as: 
%
\begin{subequations}
\begin{align}
-\nabla p + \mu \triangle \mbu = 0 \quad\text{in}\quad
  \RR^2\setminus\gamma  &  \\
% 
\nabla \cdot \mbu = 0  \quad\text{in}\quad \RR^2\setminus\gamma &\\
%
\mbu(\mbx) \rightarrow \mbu_\infty(\mbx) \quad \text{as} \quad  \norm{\mbx} \rightarrow \infty & 
 \end{align} \label{eqn:governing}
\end{subequations}%
%
The modified jump conditions in the case of a semi-permeable vesicle can be summarized as follows.  Let $\mbn$ the outward normal to $\gamma$ and $\jump{\cdot}$ denote the jump across the interface. The jump condition for the hydrodynamic stress remains the same:
\beq \jump{\Sigma\cdot\mbn} = \mbf_\text{mem}, \eeq
where $\Sigma$ is the hydrodynamic stress tensor. The fluid velocity is continuous across the membrane i.e.,
\beq \jump{\mbu} = 0. \eeq
However, the kinematic condition is modified to account for the fact that the fluid can move through the membrane. A standard choice appears to be
%
\beq \mbu - \dot{\mbx} = - \beta (\mbf_\text{mem} \cdot \mbn) \mbn \quad\text{on}\quad\gamma. \eeq
%
Here, $\beta$ is a constant and the membrane force is the usual tension + bending.

@@

\section{Integral equation formulation}
There are multiple ways to convert the PDE problem into integral
equations. Based on my derivation, the coupled set of
integro-differential equations satisfy the given interfacial conditions:
\beq 
  \dot{\mbx} = \mbu_\infty(\mbx) + \beta (\mbf_\text{mem}\cdot\mbn)\mbn + \sgl [\mbf_\text{mem}], 
  \label{eqn:vesVelocity}
\eeq
\beq 
  \mbx_s\cdot\dot{\mbx}_s = 0 
\eeq
This satisfies all the jump conditions. As usual, $\sgl$ is the single layer potential.

\section{Numerics}
To address stiffness associated with high-order derivatives,
equation~\eqref{eqn:vesVelocity} is discretized with a semi-implicit
time stepping method similar to our previous works.  First, we introduce
the notation
\begin{align}
  B[\mbx]\mbf &= -\kappa_b\frac{d^4}{ds^4} \mbf, \\
  T[\mbx]\sigma &= (\sigma \mbx_s)_s, \\
  P[\mbx]\mbf &= (\mbf \cdot \mbn) \mbn, \\
  D[\mbx]\mbf &= \mbx_s \cdot \mbf_s,
\end{align}
where $\mbn$ is the outward unit normal of the vesicle
parameterized by $\mbx$, and $d/ds$ is its arclength derivative.
Dropping the dependence of the operators on $\mbx$, the dynamics of the
vesicle are governed by
\begin{align}
  \dot{\mbx} &= \mbu_{\infty}(\mbx) + 
  \beta P(B\mbx + T\sigma) + \sgl(B\mbx + T\sigma), \\
  D \dot{\mbx} &= 0.
\end{align}
Using the notation $B^N$ to denote the bending operator due to the
vesicle configuration at time $t_N$, and similar notation for the other
operators, we apply the semi-implicit time stepping method
\begin{align}  
  \frac{\mbx^{N+1} - \mbx^N}{\Delta t} = \mbu_\infty(\mbx^N) 
  + \beta P^N(B^N\mbx^{N+1} + T^N\sigma^{N+1}) 
  + \sgl^N(B^N\mbx^{N+1} + T^N\sigma^{N+1}).
\end{align}
Applying $D$ to the first-order time discretization of
equation~\eqref{eqn:vesVelocity}, the inextensibility condition is
\begin{align*}
  &D^N(\mbx^{N+1} - \mbx^N) = 0 \\ 
  \Rightarrow &D^N\mbx^{N+1} = D^N \mbx^N \\ 
  \Rightarrow &D^N\mbx^{N+1} = 1.
\end{align*}
Finally, we write the time stepping method in matrix form as
\begin{align}
  A \left(
    \begin{array}{c}
      \mbx^{N+1} \\ \sigma^{N+1}
    \end{array}
  \right) = 
  \left(
    \begin{array}{c}
      \mbx^{N} + \Delta t \mbu_\infty(\mbx^N) \\ \mathbf{1}
    \end{array}
  \right),
\end{align}
where
\begin{align}
 A = \left(
  \begin{array}{cc}
    I + \Delta t \beta P^N B^N + \Delta t \sgl^N B^N & 
    -\Delta t P^N T^N - \Delta t \sgl^N T^N \\
    D^N & 0
  \end{array}
  \right).
\end{align}

>>>>

\documentclass[aps,prl,showpacs]{revtex4}
\usepackage{amsmath,amsfonts,bm, color}
\usepackage{graphicx,epsfig}
\usepackage{todonotes}
%Dfig puts a figure of specified size at the place indicated.
\newcommand{\Dfig}[2]{\epsfig{figure=#1,width=#2cm}}

@@

\title{Semi-permeable vesicles: mathematical formulation, numerics and physics}
\author{Ashley Gannon, Bryan Quaife, Shravan Veerapaneni, and Yuan-Nan Young}

@@

\section{Problem formulation} \label{sc:formulate}
Consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid domain, subjected to an imposed flow
$\mbu_\infty(\mbx)$, for any $\mbx \in \RR^2$.  Assume that the interior
and exterior fluids have the same viscosity $\mu$. Let $\mbx$ be the
position of the interface $\gamma$, $\mbu$ be the fluid velocity, and
$p$ be the pressure. In the vanishing Reynolds number limit, the
governing equations for the ambient fluid are
%
\begin{subequations}
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \mbu &= 0, 
  &&\mbx \in \RR^2 \setminus \gamma, \\
% 
  \nabla \cdot \mbu &= 0,  &&\mbx \in \RR^2 \setminus \gamma, \\
%
  \mbu(\mbx) &\rightarrow \mbu_\infty(\mbx),
    \qquad &&\norm{\mbx} \rightarrow \infty.
\end{alignat}
\end{subequations}%
%
Letting $\mbn$ be the outward normal to $\gamma$, $\jump{\cdot}$ denote
the jump across the interface, and $T$ denote the hydrodynamic stress
tensor, the boundary conditions along the vesicle are
\begin{align}
  \label{eqn:jump1}
  \jump{T\cdot\mbn} &= \mbf_\text{mem}, \\
  \label{eqn:jump2}
  \jump{\mbu} &= 0,
\end{align}
where $\mbf_\text{mem}$ is the standard membrane force that includes a
bending and a tension term. In particular, $\mbf_\text{mem} =
\mbf_\text{ben} + \mbf_\text{ten}$, where
\begin{align}
  \label{eqn:membraneForces}
  \mbf_\text{ben} = -\kappa_b \mbx_{ssss}, \qquad
  \mbf_\text{ten} = (\sigma \mbx_s)_s,
\end{align}
where $\kappa_b$ is the bending stiffness and $\sigma$ is a Lagrange
multiplier, which acts like the tension, which is used to satisfy the
inextensibility condition.

The kinematic condition is modified to account for the fact that the
fluid can move through the membrane. A standard choice appears
is~\cite{}
\begin{align}
  \mbu - \dot{\mbx} = - \beta (\mbf_\text{mem} \cdot \mbn) \mbn, \qquad
  \mbx \in \gamma,
\end{align}
where $\beta$ is a semi-permeability rate.

\subsection{Integral equation formulation}
Since the flow is governed by the linear Stokes equations, it can be
represented in terms of layer potentials. Given the boundary
conditions~\eqref{eqn:jump1} and~\eqref{eqn:jump2} required on the
interface, the fluid velocity can be represented as the single-layer
potential
\begin{align}
  \mbu(\mbx) = \mbu_\infty(\mbx) + \sgl[\mbf_\text{mem}](\mbx), \quad
    \mbx \in \Omega,
\end{align}
where
\begin{align}
  \sgl[\mbf](\mbx) = \frac{1}{4\pi\mu} \int_{\gamma} \left(
    -\mathbf{I} \log\rho + \frac{\mbr \otimes \mbr}{\rho^2} \right)
    \mbf(\mby) ds_{\mby},
\end{align}
where $\mbr = \mbx - \mby$ and $\rho = \norm{\mbr}$. Then, by the
kinematic equation, the dynamics of the vesicle 
\begin{align}
  \label{eqn:vesVelocity}
  \frac{d\mbx}{dt} = \mbu_\infty(\mbx) + \beta (\mbf_\text{mem}\cdot\mbn)\mbn
  + \sgl [\mbf_\text{mem}](\mbx).
\end{align}
Finally, the inextensibility condition is given by
\begin{align}
  \mbx_s\cdot\dot{\mbx}_s = 0.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle membrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\mbx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\mbx \cdot \mbn)\, ds = 
    \frac{1}{2}\int_{0}^{2\pi} \left(\mbx \cdot \pderiv{\mbx}{\theta}^\perp\right)
    \, d\theta.
\end{align}

\todo[inline]{Can do this part of the calculation using FTC}
Taking the time derivative of $A(t)$,
\begin{align}
  \dot{A}(t) &= \frac{1}{2}\int_{0}^{2\pi} \left(\dot{\mbx} \cdot
  \pderiv{\mbx}{\theta}^{\perp} + 
    \mbx \cdot \pderiv{\dot{\mbx}}{\theta}^{\perp}\right) d\theta, \\
  &= \frac{1}{2}
  \int_{0}^{2\pi}(\dot{\mbx}\cdot\mbn)\left\|\pderiv{\mbx}{\theta}\right\|d\theta +
     \frac{1}{2}\int_{0}^{2\pi}
    \left(\mbx \cdot \pderiv{\dot{\mbx}}{\theta}^{\perp}\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\mbf\cdot\mbn\right)ds +
  \frac{1}{2}\int_{0}^{2\pi}\left(\mbx \cdot \pderiv{}{t}\left(\mbn
  \left\|\pderiv{\mbx}{\theta}\right\|\right)\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\mbf\cdot\mbn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\mbx\cdot\dot{\mbn}\right)ds + \frac{1}{2}\int_{0}^{2\pi}
  \frac{\mbx\cdot\mbn}{\left\|\pderiv{\mbx}{\theta}\right\|}
  \left(\pderiv{\mbx}{\theta}\cdot
  \pderiv{\dot{\mbx}}{\theta}\right)d\theta \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\mbf\cdot\mbn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\mbx\cdot\dot{\mbn}\right)ds + \frac{1}{2}\int_{\gamma}
  (\mbx\cdot\mbn) (\mbx_s \cdot \dot{\mbx}_s)\, ds.
\end{align}
To satisfy the inextensibilty condition, $\mbx_s \cdot \dot{\mbx}_s = 0$.
Therefore the time derivative of $A(t)$ is
\begin{align}
     \dot{A}(t) &= \frac{\beta}{2} \int_{\gamma}\left(\mbf\cdot\mbn\right)ds
     + \frac{1}{2} \int_{\gamma}\left(\mbx\cdot\dot{\mbn}\right)ds.
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\mbf\cdot\mbn\right)ds-\frac{1}{2}\int_{\gamma}\left(\mbx^{\perp}\cdot\dot{\mbx}_s\right) ds.
\end{align}
Applying integration by parts,
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\mbf\cdot\mbn\right)ds+\frac{1}{2}\int_{\gamma}\left(\mbx^{\perp}_s\cdot\dot{\mbx}\right) ds,
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\mbf\cdot\mbn\right)ds+\frac{1}{2}\int_{\gamma}\left(\dot{\mbx}\cdot\mbn\right),\\
    \dot{A}(t) &= \beta \int_{\gamma}\left(\mbf\cdot\mbn\right)ds.
\end{align}
At steady state,
\begin{align}
    \dot{A}(t) &= \beta \int_{\gamma}\left(\mbf\cdot\mbn\right)ds =0.
\end{align}
We now replace $\mbf$ with its definition and apply integration by parts
\begin{align}
   0=\beta\int_{\gamma} 
    (-\kappa_b\mbx_{ssss}+(\sigma\mbx_s)_s)\cdot\mbn\, ds =
   \beta\int_{\gamma}\left(\kappa_b\mbx_{sss}-\sigma\mbx_s\right)\cdot
     \mbn_s\, ds
\end{align}
From the Frenet-Serret formulas, we have 
\begin{align}
     -\beta\int_{\gamma}
     \left(\kappa_b\mbx_{sss}-\sigma\mbx_s\right)\cdot\kappa\mbx_s ds &=0.
\end{align}

Applying integration by parts and Frenet-Serret again
\begin{align}
     \beta\int_{\gamma}
     \kappa_b\kappa\left(\mbx_{ss}\cdot\mbx_{ss}\right)ds + \beta
     \int_{\gamma} \kappa_b \kappa_s(\mbx_{ss} \cdot \mbx_s)ds +\beta\int_{\gamma}\kappa\sigma ds  &= 0,\\
    \beta\int_{\gamma} \kappa_b\kappa(\kappa\mbn \cdot
     \kappa\mbn)+ \beta \int_{\gamma} \kappa_b \kappa_s(\mbx_{ss} \cdot
     \mbx_s)ds +\beta\int_{\gamma}\kappa\sigma ds &= 0,
\end{align}
which we re-write as
\begin{align}
   \beta\int_{\gamma}\kappa_b\kappa^3 ds + \beta \int_{\gamma} \kappa_b
   \kappa_s(\mbx_{ss} \cdot \mbx_s)ds +\beta\int_{\gamma} \kappa\sigma ds = 0.
\end{align}
Applying Frenet-serret again,
\begin{align}
    \beta\int_{\gamma}\kappa_b\kappa^3 ds + \beta \int_{\gamma} \kappa_b
    \kappa_s(\kappa\mbn \cdot \TT)ds +\beta\int_{\gamma} \kappa\sigma ds = 0, \\
    \beta\int_{\gamma}\kappa_b\kappa^3 ds +\beta\int_{\gamma} \kappa\sigma ds  = 0.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\subsection{Discretization in Space}

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:membraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\mbx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\mbx]\mbf &= -\frac{d^4}{ds^4} \mbf, \\
  T[\mbx]\sigma &= (\sigma \mbx_s)_s, \\
  D[\mbx]\mbf &= \mbx_s \cdot \mbf_s, \\
  P[\mbx]\mbf &= (\mbf \cdot \mbn) \mbn. 
\end{align}
Here, the normal vector $\mbn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\mbf$ to mean $B[\mbx]\mbf$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\mbx}{dt} &= \mbu_{\infty}(\mbx) + 
  \beta P(B\mbx + T\sigma) + \sgl(B\mbx + T\sigma), \\
  D \frac{d\mbx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t_N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\mbx^{N+1} - \mbx^N}{\Delta t} = \mbu_\infty(\mbx^N) 
  + \beta P^N\left(B^N\mbx^{N+1} + T^N\sigma^{N+1}\right) 
  + \sgl^N\left(B^N\mbx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\mbx^{N+1} = 1.
\end{align}

\todo[inline]{A few lines about SDC}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Examples}
We consider a single semi-permeable vesicle in four different flow
regimes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quiescent Flow}
\begin{itemize}
  \item Always goes to a circle
  \item Area formula works
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Shear Flow}
\begin{itemize}
  \item Never goes to a circle
  \item Area formula works
  \item Mild dependence on initial shape
  \item Most slender tank treading shapes seen for small $\beta$ and
    large $\chi$
\end{itemize}

%\input{figures/shearPhaseDiagram.tikz}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Extensional Flow}
\begin{itemize}
  \item ???
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Poiseuille Flow}
\begin{itemize}
  \item ???
\end{itemize}

 \newpage
 \paragraph{\bf \Huge OLD STUFF}

@@

%%%%%%%%%%%%%%
Finally, we write the time stepping method in matrix form as
\begin{align}
  A \left(
    \begin{array}{c}
      \mbx^{N+1} \\ \sigma^{N+1}
    \end{array}
  \right) = 
  \left(
    \begin{array}{c}
      \mbx^{N} + \Delta t \mbu_\infty(\mbx^N) \\ \mathbf{1}
    \end{array}
  \right),
\end{align}
where
\begin{align}
 A = \left(
  \begin{array}{cc}
    I + \Delta t \beta P^N B^N + \Delta t \sgl^N B^N & 
    -\Delta t P^N T^N - \Delta t \sgl^N T^N \\
    D^N & 0
  \end{array}
  \right).
\end{align}
